---
title: "AE 03 - Visualizing Star Wars Characters"
author: "Dina Alarcon"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: show
  pdf_document:
    toc: true
    latex_engine: xelatex
    keep_tex: false
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

## Introduction

In this application exercise, you'll practice creating various types of visualizations using the Star Wars characters dataset. This dataset contains information about characters from the Star Wars universe, including their physical characteristics, homeworld, and species.

**In this application exercise, you will:**

- Explore data with `glimpse()`
- Create scatterplots with multiple aesthetics
- Make histograms with appropriate binwidths
- Create visualizations for different variable types
- Interpret relationships between variables
- Customize plot labels and aesthetics
- Choose appropriate plot types for different data types

**Estimated time:** 30-45 minutes

---

## Getting Started

### Fork and Clone Workflow

**Quick reminder:** You should be working in YOUR forked `application-exercises` repository.

1. Navigate to the `ae-03-starwars` folder in YOUR fork
2. Open `ae-03-starwars.Rmd`
3. Update the YAML with your name
4. Click **Knit** to verify it works

**If you need detailed fork-and-clone instructions, see AE 01.**

---

## Packages
```{r load-packages, include=FALSE}
library(tidyverse)
```

**What this package does:**
- **tidyverse:** Includes ggplot2 for visualization and dplyr for data wrangling

---

## The Data

The `starwars` dataset comes built-in with the tidyverse. It contains information on Star Wars characters from the first seven films.

---

## Exercise 1: Glimpse at the Data

Glimpse at the `starwars` data frame to understand its structure.
```{r glimpse-starwars}
glimpse(starwars)
```

**Answer the following questions:**

**How many observations (rows) are there?** There are 87 observation rows. 

**How many variables (columns) are there?** There are 14 variable columns.

**Name three numerical variables:**

1. Height
2. Mass
3. birth_year

**Name three categorical variables:**

1. Name
2. hair_color
3. skin_color

**What does each observation represent?**

Each observation represent one Star Wars character. 

---

## Exercise 2: Understanding Aesthetics

The plot below maps several variables to aesthetics. However, there's an issue with how color is being used.

**Original code:**
```{r scatterplot-original, eval=FALSE}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "pink")
```

**Task 2.1:** Run the code above (in your Console, not by knitting). What happened to the `color = gender` mapping?

**Your answer:**

The color = gender mapping disappeared, and all the points became pink.

**Task 2.2:** Why did this happen?

**Hint:** Think about where `color = "pink"` appears vs. where `color = gender` appears.

**Your answer:**

This happened because color = "pink" is set inside geom_point(), which overrides the color = gender mapping inside aes().

**The issue:** When you set `color = "pink"` inside `geom_point()`, it overrides the aesthetic mapping `color = gender` from `aes()`. A fixed color set inside a `geom_*()` function always wins over a mapped color set inside `aes()` — you can't have both at once.

**Task 2.3:** Since you can't keep the `color = gender` mapping *and* make every point pink, edit the code below to drop the conflicting mapping while keeping the other aesthetic mappings (`x`, `y`, and `size`) intact:
```{r scatterplot-pink}
ggplot(starwars, 
       aes(x = height, y = mass, size = birth_year)) +
  geom_point(color = "pink")
```

**What your finished plot should show:**

- x-axis: Height of characters
- y-axis: Mass (weight) of characters  
- Size of points: Birth year (larger = born later)
- All points are pink

---

## Exercise 3: Adding Labels

Add appropriate labels for the title, x-axis, y-axis, and size legend. Uncomment the lines and fill in the blanks.
```{r scatterplot-labels}
ggplot(starwars, 
       aes(x = height, y = mass, color = gender, size = birth_year)) +
  geom_point(color = "#30509C") +
  labs(
    title = "Height and Mass of Star Wars Characters",           # Title about relationship between height and mass
    x = "Height (cm)",               # What is on the x-axis?
    y = "Mass (kg)",               # What is on the y-axis?
    size = "Birth Year"             # What does size represent?
  )
```

**Hints:**

- Title should describe what the plot shows
- x and y labels should include units if relevant (height in cm, mass in kg)
- size label describes what the size of points represents

**Task 3.1:** Why is there no `color` label in the `labs()` function?

**Your answer:**

There is no color label because color is set to one fixed color inside geom_point().The plot does not need a color legend, because its being mapped to a variable. 

---

## Exercise 4: Create a Histogram

Pick a single numerical variable and make a histogram of it. Select a reasonable binwidth.

**First, explore the variable:**
```{r explore-for-histogram}
# Choose a variable and look at its range
starwars %>%
  summarise(
    min_height = min(height, na.rm = TRUE),
    max_height = max(height, na.rm = TRUE),
    mean_height = mean(height, na.rm = TRUE)
  )
```

**Note:** This example explores `height`. If you pick a different variable (like `mass` or `birth_year`) for your histogram below, swap it into the code above too and rename the summary columns to match (e.g., `min_mass`, `max_mass`, `mean_mass`), so your exploration actually matches what you're plotting.

**Now create your histogram:**
```{r histogram, eval = FALSE}
# Fill in the blanks
ggplot(starwars, aes(x = height)) +
  geom_histogram(binwidth = 10, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Star Wars Character Heights",
    x = "Height (cm)",
    y = "Number of Characters"
  )
```

**Hints:**

- Choose from: height, mass, birth_year
- Try different binwidths: 10, 20, 50 (depending on your variable)
- Use `fill` to color the bars, `color` to color the borders

**Task 4.1:** What binwidth did you choose and why?

**Your answer:**

I chose a binwidth of 10 centimeters because it helps separate the heights into clear groups. Choosing 10 cm helps not create too many and/or too few bars. 

**Task 4.2:** Describe the distribution. Is it symmetric, skewed left, or skewed right? Are there any outliers?

**Your answer:**

The distribution is slightly skewed to the left. Most characters in the left are between approximately 160 and 200 centimeters tall. There are a few short characters below 120 centimeters. There is one taller character around 260 centimeters that appears to be considered outliers. 

---

## Exercise 5: Numerical vs. Categorical

Pick a numerical variable and a categorical variable and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Boxplot: `geom_boxplot()`
- Violin plot: `geom_violin()`
- Multiple histograms: `geom_histogram()` with `facet_wrap()`

**Example combinations to try:**

- height by gender
- mass by species
- birth_year by sex
```{r num-cat, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender, y = height)) +
  geom_boxplot(fill = "lightblue") +
  labs(
    title = "Height of Star Wars Characters by Gender",
    x = "Gender",
    y = "Height (cm)"
  )
```

**Interpretation:**

What does this visualization tell you about the relationship between these two variables?

The visualization tells me about the relationship between these two variable is that masculine characters generally have a higher median height than feminine characters. They also have wider range of heights and more outliers, including several unusually short and tall characters. Characters with missing gender information have heights near the middle of the other two groups. 

---

## Exercise 6: Bar Plot of Categorical Variable

Pick a single categorical variable from the dataset and make a bar plot of its distribution.

**Suggested variables:**

- gender
- sex  
- species
- eye_color
- hair_color
```{r barplot, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Number of Star Wars Characters by Gender",
    x = "Gender",
    y = "Number of Characters"
  )
```

**Task 6.1:** Which category has the most observations?

**Your answer:** The category that has the most observations is masculine category with approximately 66 characters.

**Task 6.2:** If your plot has many categories with small bars that are hard to read, try flipping the coordinates:
```{r barplot-flipped, eval = TRUE}
# Same code as above, but add coord_flip() at the end
ggplot(starwars, aes(x = gender)) +
  geom_bar(fill = "coral") +
  labs(
    title = "Number of Star Wars Characters by Gender",
    x = "Gender",
    y = "Number of Characters"
  ) +
  coord_flip()
```

---

## Exercise 7: Two Categorical Variables

Pick two categorical variables and make a visualization to visualize the relationship between the two variables.

**Suggested plot types:**

- Stacked bar plot: `geom_bar()`
- Grouped bar plot: `geom_bar(position = "dodge")`
- Faceted bar plot: `geom_bar()` with `facet_wrap()`
- Heatmap: `geom_tile()` with counts

**Suggested combinations:**

- gender and species
- sex and eye_color
- gender and hair_color

**Option 1: Stacked bar plot**
```{r cat-cat-stacked, eval = TRUE}
# Your code here
ggplot(starwars, aes(x = gender, fill = hair_color)) +
  geom_bar() +
  labs(
    title = "Hair Color of Star Wars Characters by Gender",
    x = "Gender",
    y = "Count",
    fill = "Hair Color"
  )
```

**Option 2: Grouped bar plot (side-by-side)**
```{r cat-cat-grouped, eval = FALSE}
# Your code here
ggplot(starwars, aes(x = ___, fill = ___)) +
  geom_bar(position = "dodge") +
  labs(
    title = "___",
    x = "___",
    y = "Count",
    fill = "___"
  )
```

**Choose one of the above and interpret:**

What does this visualization tell you about the relationship between these two variables?

The relationship between these two variables is that masculine characters make up the largest group and have the greatest variety of hair colors. “None” is the most common hair-color category among masculine characters, while feminine characters have fewer observations and a smaller variety of hair colors.

---

## Exercise 8: Multi-Variable Visualization

Pick two numerical variables and two categorical variables and make a visualization that incorporates all of them.

**Suggested approach:**

- x-axis: numerical variable 1
- y-axis: numerical variable 2
- color: categorical variable 1
- facets: categorical variable 2

**Example combination:**

- height (numerical)
- mass (numerical)
- gender (categorical - use for color)
- species (categorical - use for facets, maybe filter to top 3-4 species)

**Starter template:**
```{r multi, eval=TRUE}
# You might want to filter first to avoid too many facets
starwars_filtered <- starwars %>%
  filter(species %in% c("Human", "Droid", "Wookiee"))

#plot
ggplot(starwars_filtered,
       aes(x = height, y = mass, color = gender)) +
  geom_point(size = 3) +
  facet_wrap(~ species) +
  labs(
    title = "Height and Mass by Species and Gender",
    x = "Height (cm)",
    y = "Mass (kg)",
    color = "Gender"
  )
```

**Interpretation:**

What patterns do you notice? How do the relationships differ across the facets?

For the human characters, taller characters generally weigh more. The droids have very different heights and weights. Both Wookiees are tall and heavy, and the taller Wookiee also weighs more. Most of the characters shown are masculine.

---

## Exercise 9: Reflection

**Task 9.1:** Which type of plot was most useful for exploring this data? Why?

**Your answer:**

I think the scatterplot was the most useful because it made the relationship between height and weight easy to see. The colors and separate sections also made it easier to compare gender and species.

**Task 9.2:** What was the most interesting pattern or relationship you found?

**Your answer:**

The most interesting pattern I found was that taller characters usually weighed more, especially the human characters. I also noticed that the droids had very different heights and weights.

**Task 9.3:** What did you find most challenging about creating these visualizations?

**Your answer:**

The most challenging part was choosing the right variables and graph for each exercise. I also had to make sure that every graph had a title and the correct labels.
---

## Knit and Submit

**Before you finish:**

1. ✅ Make sure all code chunks run without errors
2. ✅ Fill in all answer spaces and interpretations
3. ✅ Knit to HTML first to verify everything works
4. ✅ Review the HTML output - check that all plots display correctly

**Git workflow:**

1. Click the **Git** pane
2. Check boxes next to all changed files
3. Click **Commit**
4. Write a commit message: "Completed AE 03 - Star Wars"
5. Click **Commit**
6. Click **Push**
7. Verify your work is on GitHub in **YOUR fork**

---

## Generate PDF for Canvas Submission

Once everything works in HTML:

1. Click the **Knit** dropdown arrow
2. Select **Knit to PDF**
3. Wait for PDF generation
4. Find `ae-03-starwars.pdf` in the Files pane
5. Export and save to your computer
6. Submit the PDF to Canvas

**Remember:** Submit to BOTH GitHub (your .Rmd) AND Canvas (the PDF). Grading is based on the PDF, so double-check it before submitting — but your GitHub repo should be up to date too.

---

## Troubleshooting

**Common issues:**

- **Error: "object not found":** Make sure you loaded the tidyverse package
- **Plot doesn't show up:** Make sure the code chunk is set to run (not `eval: false`)
- **Too many categories in plot:** Filter the data first or use `coord_flip()` for better readability
- **NAs in plot:** This is normal for this dataset - some characters have missing values

**Aesthetic mapping issues:**

- **Color not showing:** Make sure `color = variable` is inside `aes()`, not `geom_point()`
- **Want to set all points to one color:** Put `color = "colorname"` OUTSIDE `aes()`, inside `geom_point()`

**Plot type selection:**

- **One numerical variable:** histogram or density plot
- **One categorical variable:** bar plot
- **Numerical vs. categorical:** boxplot, violin plot, or faceted histogram
- **Two categorical:** stacked/grouped bar plot or faceted plot
- **Two numerical:** scatterplot
- **Multiple variables:** use color, size, shape, facets

**PDF issues:**

- **LaTeX not found:** Install tinytex: `tinytex::install_tinytex()`
- **Plots too large for PDF:** Adjust `fig.width` and `fig.height` in chunk options

**If you're stuck:**

- Review the walkthrough video
- Check ggplot2 documentation: `?geom_point`, `?geom_histogram`, etc.
- Ask on the discussion board
- Attend office hours

---

## Bonus: Quick Reference

**Basic ggplot structure:**
```r
ggplot(data, aes(x = var1, y = var2)) +
  geom_*() +
  labs(title = "...", x = "...", y = "...")
```

**Common geoms:**

- `geom_point()`: scatterplot
- `geom_histogram()`: histogram
- `geom_bar()`: bar plot
- `geom_boxplot()`: boxplot
- `geom_violin()`: violin plot
- `geom_line()`: line plot

**Common aesthetics (inside `aes()`):**

- `x = variable`: x-axis
- `y = variable`: y-axis
- `color = variable`: color by variable
- `fill = variable`: fill color by variable
- `size = variable`: size by variable
- `shape = variable`: shape by variable

**Fixed aesthetics (outside `aes()`):**

- `color = "blue"`: all points blue
- `fill = "red"`: all filled red
- `size = 3`: all points size 3
- `alpha = 0.5`: transparency (0 = transparent, 1 = opaque)
